\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-09-24 chapters/bci.tex
Intro, P300 stimulation

Independent home use of a brain-computer interface by people with
amyotrophic lateral sclerosis.
\todo{Shortly describe challenges from the patient point of view, related to
this thesis}
\todo{do something with Wolpaw (2018) Independent home use of a brain-computer interface by people with
amyotrophic lateral sclerosis.}

\todo{
Locked-in -\> communicate
Paraplegic -\> walk again
tetraplegic, locked-in syndrome
Regaining capabilities and improving quality of life
Start from wolpaw examples, for spelling
Cite some catchy state of the art examples (walk again, speech avatar Metzger, handwriting
willet, neuralink)
}
\todo{put focus on communication BCI}
\section{A direct interface to the brain}
\comment{This section describes a bci in layman's terms, without too much
scientific background, jargon and literature, in the following sections of this
chapter a more scientific style is used}
A \ac{bci} reads your mind.
\ac{bci}s are an entertaining subject with many futuristic and more established
applications.
Devices processing direct inputs from the central nervous system can be
useful for rehabilitation and medical diagnosis or treatment.
They have always been especially promising as assistive technology for
paralyzed patients limited in their communication ability. \todo{expand more on this last
sentence, put patients first in introduction, see above}
Furthermore, they bring a new paradigm to human-computer interaction,
especially when paired with virtual or augmented reality.

@@

Finally, an important concept in the evaluation of a \ac{bci} decoder, and of \acp{bci}
in general, is \ac{itr}.
\Ac{itr} reflects how fluently a \ac{bci} can be used for communication and can
be calculated as
\begin{equation}
	\text{ITR} = Q\left(\log_2N+P\log_2P+(1-P)\log_2\frac{1-P}{N-1}\right)
\end{equation}
\Ac{itr} is expressed in bits/s and is dependent on $N$, the number of
different options that can be selected per trial, $P$, the selection accuracy
of the decoder, and $Q$, the number of trials per second.
The parameters $N$, $P$ and $Q$ of this formula give us some insight in the
building blocks a successful, high-\ac{itr} \ac{bci}.
To improve performance, we can aim to
\todo{enumerate}
\begin{enumerate*}[label={\arabic*})]
\item increase $N$ by selecting a paradigm and interface that offer a
broad range of informative selections per trial,
\item increase $P$ by engineering more performant machine learning methods for
decoding, and
\item increase $Q$ by selecting a paradigm and allows fast stimulation or
responses.
\end{enumerate*}

\section{A case study: the visual oddball paradigm}

The visual oddball paradigm is a reactive, stimulus-dependent, synchronous
\ac{bci} paradigm.
good choice because it scores high on the parameters above
well established, good research, first application cite
Flashing in various configurations, such as row-colum or in groups,

\subsection{Stimulation}
comfortable, preferred over SSVEP, similar ITR is this true?
\todo{figure of P3 matrix}
\todo{cite P300 literature, first example and current sota}

@@

@@

@@

@@

\subsection{Preprocessing and feature extraction}
referencing, filtering, epoching, baselining
\subsection{Decoding ERPs}
tLDA, beamforming
Linear decoding techniques

>>>>

\todo{comments liesel paradigms}
People suffering from heavy paralysis or disability due to
neurodegenenerative diseases, stroke, traumatic brain or spine injury
become entirely reliant on caregivers or family members in their day to day
routine.
Wouldn't it be great if we could help them regain some quality of life by
making them more independent?

The most basic ability any independent human being needs, is interaction with
their environment.
In it's core essence, this is the ability to communicate your intentions,
emotions, frustrations or thoughts to someone or something, which then might
act upon them.
For an able-bodied person, physical interaction happens through body movements
and communication is usually done through speech and body language, which
both require sending signals through the body's nervous system to control
muscles.
Not everyone has this capability.

The key problem for heavily paralyzed patients is that the mind wants to go
where the body can't follow.
Couldn't we therefore design a system that directly interacts with the mind?
This system can then form an interface between the mind and any kind of
technology, like a robot arm or speech synthesizer, interacting with the world
on the user's behalf.
If the mind's seat is in the brain, and we want to interface this to a digital
control process, such a device becomes the \ac{bci}\footnote{Sometimes also
termed brain-machine interface (BMI) when coupled to a physical actuator, like
a robot arm or an exoskeleton. The term \ac{bci} is usually preferred for
assistive technology and communication devices.}.
A \ac{bci} directly couples actions in the `real world' to the mental state of
the user.
A \ac{bci} reads your mind.

Starting in the 1970's from methods to establish minimal brain activity
(yes/no, left/right, ...)~\cite{Wolpaw2002}, \acp{bci} have gradually evolved
into sophisticated interaction schemes that restore or enhance many capabilities
by bypassing part of the nervous system and muscle control.
By now, classic \acp{bci} methods have matured outside the lab setting and can be
used as assistive communication technology by patients with impaired
speech~\cite{Wolpaw2018}.
Impressive cutting edge examples of experimental \acp{bci} include a brain-spine
interface allowing a paraplegic patient to walk again\cite{Lorach2023},
a \ac{bci} translating internal speech to a facial avatar speaking the imagined
text~\cite{Metzger2023}, and fast communication through decoding imagined
handwritten symbols~\cite{Willett2021}.
Recently, the idea has also gained popular traction through Elon Musk's
Neuralink~\cite{Musk2019} initiative, which imagines a multipurpose \ac{bci}.

\todo{put focus on communication BCI}

\section{A direct interface to the brain}
\comment{This section describes a bci in layman's terms, without too much
scientific background, jargon and literature, in the following sections of this
chapter a more scientific style is used}
\ac{bci}s are an entertaining subject with many futuristic and more established
applications.
Devices processing direct inputs from the central nervous system can be
useful for rehabilitation and medical diagnosis or treatment, but
have always been especially promising as assistive technology for
paralyzed patients limited in their communication ability.
Furthermore, they bring a new paradigm to human-computer interaction,
especially when paired with virtual or augmented reality.

@@

Finally, an important concept in the evaluation of a \ac{bci} decoder, and of \acp{bci}
in general, is \ac{itr}.
\Ac{itr} reflects how fluently a \ac{bci} can be used for communication and can
be calculated as
\begin{equation}
	\text{ITR} = Q\left(\log_2N+P\log_2P+(1-P)\log_2\frac{1-P}{N-1}\right)
\end{equation}
\Ac{itr} is expressed in bits/s and is dependent on $N$, the number of
different options that can be selected per trial, $P$, the selection accuracy
of the decoder, and $Q$, the number of trials per second.
The parameters $N$, $P$ and $Q$ of this formula give us some insight in the
building blocks a successful, high-\ac{itr} \ac{bci}.
To improve performance, we can aim to
\begin{enumerate*}[label={\arabic*})]
\item increase $N$ by selecting a paradigm and interface that offer a
broad range of informative selections per trial,
\item increase $P$ by engineering more performant machine learning methods for
decoding, and
\item increase $Q$ by selecting a paradigm and allows fast stimulation or
responses.
\end{enumerate*}

\section{A case study: the visual oddball \ac{bci}}

The visual oddball paradigm is a \emph{reactive}, \emph{stimulus-dependent},
\emph{synchronous} \ac{bci} paradigm with all the benefits and drawbacks
Nevertheless, it can score relatively high on the \ac{itr} parameters established above.
The brain response of interest is the visual \ac{erp}, which can be accurately
measured and decoded from non-invasive \ac{eeg} signals with a relatively low
computational effort and short calibration time.
First established by~\cite{Farwell1988} in 1988, it well supported by literature
and has been proven to work for patients in day-to-day home
use~\cite{Wolpaw2018}.

\subsection{Stimulation paradigm}
One by one, visual elements on a computer screen are \emph{intensified} for a
short period of time by changing in color, brightness or size.
On each flash observed by the user, specific brain activity is elicited.
If one of these intensified elements is rare with respect to the others, i.e. it is the
\emph{odd-one out} or \emph{oddball}, this brain activity is altered.
A stimulus can be rare either due to its inherent properties like color or
location
Crucially, this is also the case if it is `marked' as rare by the user, e.g.
by paying explicit attention to only a given stimulus and not the others or
counting how many times it flashes.
If these visual elements represent letters, and we know the timing of the
intensifications of each letter, we can now establish communication by detecting
for which letter an oddball response was present in the brain signal at its
time of intensification.

\Ac{itr} in the oddball paradigm is optimized by intensifying
multiple targets at once in a row-collumnn strategy and using a sequence of
selections, giving rise to the classic
P300 matrix speller of~\cite{Farwell1988}.
Other optimizations increase response \ac{snr}, like using flashing face
images as intensifications~\cite{Jin2012} or adding distinct colors and shapes to the
stimuli~\cite{Treder2011};

Sometimes Flashing in various configurations, such as row-column or in groups,
other flashing patterns
faces, feature attention
visual hex-o-spell

comfortable, preferred over SSVEP, similar ITR is this true?
\todo{figure of P3 matrix}
\todo{cite P300 literature, first example and current sota}

@@

later compoentns P3 more attention/cognitive more centrally in the brain

@@

mention VEPs more from visual system occipitally

@@

which components, at what amplitude and latency they are evoked varies per
person and depends on the type of intensification

@@

\subsection{Preprocessing and feature extraction}
referencing, filtering, epoching, baselining
\subsection{Decoding ERPs}

Usually, snr low and multiple trials necessary

tLDA, beamforming
Linear decoding techniques

### 2024-09-24 chapters/wcble.tex
Intro, P300 stimulation

@@

\section{Results}
simulation study

\subsection{Data simulation}
To verify our approach, we simulated 32 channel \ac{eeg} data with standard
10-20 electrode positions using the \texttt{simulate\_evoked} method of the
MNE (version 1.8)~\cite{Gramfort2013} software package.
\Acp{erp} were generated by projecting a source time course in a dipole in the
left hemisphere to the scalp electrodes using a boundary element method forward
model~\cite{Mosher1999} MNE's \texttt{fsaverage} source space and anatomy.
We refer to~\textcite{Gramfort2014} for implementation details.
Pink spatial and  temporal noise are then added.
Temporal noise is generated by passing gaussian noise through an infinite
impulse response filter with transfer function
$\frac{B(z)}{A(z)} = \frac{1}{1 -1z^{-1}+0.15z^{-2}}$.
For pink spatial noise, a noise covariance matrix is constructed from the
cosine distances between electrode positions and scaling the resulting spatial
frequencies by $1/f$.

@@

figure

\subsection{Latency estimation}
cble with prob/median
wcble with prob+median
cble with max
wcble with max

@@

\section{Discussion}
\todo{mention cross validation}
\todo{remove lcmv}
In this work, we introduce a new latency method that leverages data-driven,
joint spatiotemporal filtering based on the LCMV beamformer~\cite{Vliet2016}. The algorithm
integrates the spatiotemporal filtering in Woody's template matching iteration
scheme and can easily be extended to multi-component settings.
Our method attempts to model the spatial and temporal activation of
each \ac{erp} component, as well as it's latency distribution, and takes into acount the
spatial and temporal correlation of the EEG background noise.

with lda, it is in the single class case equal to shifting a spatiotemporal
lcmv beamformer over the epochs and can be used to extract latencies.
multiclass:

>>>>

\todo{comments liesel}

@@

\section{Results}

We compare our proposed \ac{wcble} algorithm to \ac{cble} on both a latency
estimation task and a binary classification task with simulated data.
Four settings are evaluated to assess both the impact of the iterative Woody
procedure and the extraction of robust latency features.
These are CBLE-median and WCBLE-median, where the median of the latency
probability distribution is used as latency estimate, and CBLE-peak and
WCBLE-peak, where the maximum of the score output is used as latency estimate.

\subsection{Data simulation}
To verify our approach, we simulated 32 channel \ac{eeg} data with standard
10-20 electrode positions using the \texttt{simulate\_evoked} method of the
MNE (version 1.8)~\cite{Gramfort2013} software package.
\Acp{erp} were generated by projecting a source time course in a dipole in the
left hemisphere to the scalp electrodes using a boundary element method forward
model~\cite{Mosher1999} MNE's \texttt{fsaverage} source space and anatomy.
Pink temporal noise is generated by passing gaussian noise through an infinite
impulse response filter with transfer function
$\frac{B(z)}{A(z)} = \frac{1}{1 -1z^{-1}+0.15z^{-2}}$.
Pink spatial noise is also added, using a noise covariance matrix constructed from the
cosine distances between electrode positions and scaling it by the inverse of
the resulting spatial.
We refer to~\textcite{Gramfort2014} for implementation details.

@@

A sample of the simulated evoked data is displayed in
Figures~\ref{fig:wcble/sim-gausian}, \ref{fig:wcble/sim-morlet} and
\ref{fig:wcble/sim-sine}.
\fullpagefig{figures/wcble/simulated-gaussian.pdf}{}{fig:wcble/sim-gaussian}
\fullpagefig{figures/wcble/simulated-morlet.pdf}{}{fig:wcble/sim-morlet}
\fullpagefig{figures/wcble/simulated-sine.pdf}{}{fig:wcble/sim-sine}
\todo{Properly align plots and label axes}

\todo{figure}

\subsection{Latency estimation}

@@

\section{Discussion}

\todo{mention cross validation}
\todo{remove lcmv}

\todo{In this work, we introduce a new latency method that leverages data-driven,
joint spatiotemporal filtering based on the LCMV beamformer~\cite{Vliet2016}. The algorithm
integrates the spatiotemporal filtering in Woody's template matching iteration
scheme and can easily be extended to multi-component settings.
Our method attempts to model the spatial and temporal activation of
each \ac{erp} component, as well as it's latency distribution, and takes into acount the
spatial and temporal correlation of the EEG background noise.}

\todo{with lda, it is in the single class case equal to shifting a spatiotemporal
lcmv beamformer over the epochs and can be used to extract latencies.
multiclass:}

### 2024-09-24 main.tex
Add gaze disabilities
Add decoders
Comments Liesel
Intro, P300 stimulation

\newcommand{\comment}[1]{
  \todo[color=yellow]{#1}
}

% Bibliography
\usepackage[backend=biber]{biblatex}
\addbibresource{references.bib}
\addbibresource{publications.bib}

% Title

@@

\part{Introduction \& Literature}
\includechapter{Brain-Computer Interfaces}{chapters/bci.tex}
\includechapter{Gaze-independent visual BCIs}{chapters/gaze_independence.tex}
\todo{The role of jitter in gaze-independent ERP decoding}
%\includechapter{Problem statement \& rationale}{chapters/problem_statement.tex}

@@

\part{Experiments \& Applications}
\includechapter{Compensating jitter for gaze-independent decoding}{chapters/covert_align.tex}
\todo{Integrating gaze-tracking data}
\includechapter{Patient cases}{chapters/patients.tex}
\includechapter{Conclusion \& recommendations}{chapters/conclusion.tex}

\backmatter%
\chapter{Funding \& acknowledgments}
\chapter{Personal contribution}
\chapter{Conflict of interest}
The authors declare no conflict of interest.
\include{chapters/cv.tex}
\end{document}

>>>>

\newcommand{\comment}[1]{
  \todo[color=yellow]{#1}
}
\newcommand*\elide{\textup{[\,\dots]}}

% Bibliography
\usepackage[backend=biber, doi=false, url=false, isbn=false, eprint=false]{biblatex}
\addbibresource{references.bib}
\addbibresource{publications.bib}

% Set bibliography font size
\renewcommand*{\bibfont}{\footnotesize}

\usepackage[inline]{enumitem}

@@

\part{Introduction \& Literature}
\includechapter{Brain-computer interfaces}{chapters/bci.tex}
\includechapter{Gaze-independent visual BCIs}{chapters/gaze_independence.tex}
\todo{The role of jitter in gaze-independent ERP decoding}
%\includechapter{Problem statement \& rationale}{chapters/problem_statement.tex}

@@

\part{Experiments \& Applications}
\includechapter{Compensating jitter for gaze-independce}{chapters/covert_align.tex}
\todo{Integrating gaze-tracking data}
\includechapter{Patient cases}{chapters/patients.tex}
\includechapter{Conclusion \& recommendations}{chapters/conclusion.tex}

\backmatter%
\chapter{Funding \& acknowledgments}
\todo{my grant}\todo{marc's projects}
\chapter{Personal contribution}
\todo{Juliette, Alixe, Kristl Claeys, Edward Schietekatte, Marc, François,
Hakim, Marie,Etienne, Philip Van Damme  }
\chapter{Conflict of interest}
The authors declare no conflict of interest.
\include{chapters/cv.tex}
\end{document}
